No one-size-fits-all solution to clean GBIF

### 2020-02-25 manuscript.tex
include antonellis feedback

\PassOptionsToPackage{unicode=true}{hyperref} % options for packages loaded elsewhere
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{lmodern}
\usepackage{setspace}
\setstretch{1.3}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provides euro and other symbols
\else % if luatex or xelatex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Automated filtering of species occurrence data can identify problems, but needs to be customized},
  pdfauthor={Alexander Zizka1,2,; Fernanda Antunes Carvalho4,5,; Alice Calvente4,5,; Ana Paula Alves da Silva4,5,; Andressa Cabral,; Arthur de Souza Soares4,5,; Eduardo Calisto Tomaz4,; Fernanda Gondim Lambert Moreira4,5,; Filipe Serrano,; Jéssica Fernanda Ramos Coelho,; Mabel Rocio Baez-Lizarazo,; Mariana Ramos Fantinati4,5,; Matheus Colli-Silva,; Moabe Ferreira Fernandes,; Nathália Michelly da Cunha Santos,; Paolla G. C. Souza,; Renata Clícia Santos da Costa,; Thais Ferreira Pinto de Araújo,; Tiago Andrade Borges Santos,; Tiago Luiz Vieira,; Valéria Fonseca Vale,; Alexandre Antonelli6},
  pdfborder={0 0 0},
  breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs}
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-2}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

% set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\usepackage{caption}
\usepackage{fancyhdr}
\usepackage{booktabs}
\pagestyle{fancy}
\fancyhead[L]{Automated filter effect}
\fancyhead[LE,RO]{\thepage}
\fancyfoot{}
\usepackage{float}
\floatplacement{figure}{H}
\usepackage{lineno}
\linenumbers
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\title{Automated filtering of species occurrence data can identify problems, but needs to be customized}
\author{Alexander Zizka\textsuperscript{1,2}, \and Fernanda Antunes Carvalho\textsuperscript{4,5}, \and Alice Calvente\textsuperscript{4,5}, \and Ana Paula Alves da Silva\textsuperscript{4,5}, \and Andressa Cabral, \and Arthur de Souza Soares\textsuperscript{4,5}, \and Eduardo Calisto Tomaz\textsuperscript{4}, \and Fernanda Gondim Lambert Moreira\textsuperscript{4,5}, \and Filipe Serrano, \and Jéssica Fernanda Ramos Coelho, \and Mabel Rocio Baez-Lizarazo, \and Mariana Ramos Fantinati\textsuperscript{4,5}, \and Matheus Colli-Silva, \and Moabe Ferreira Fernandes, \and Nathália Michelly da Cunha Santos, \and Paolla G. C. Souza, \and Renata Clícia Santos da Costa, \and Thais Ferreira Pinto de Araújo, \and Tiago Andrade Borges Santos, \and Tiago Luiz Vieira, \and Valéria Fonseca Vale, \and Alexandre Antonelli\textsuperscript{6}}
\date{}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  sDiv,German Center for Integrative Biodiversity Research (iDiv), Leipzig, Germany
\item
  Naturalis Biodiversity Center, Leiden, The Netherlands
\item
  Departamento de Genética, Ecologia e Evolução, Instituto de Ciências Biológicas, Universidade Federal de Minas Gerais, Av. Antônio Carlos 6627 - Pampulha, 31270-901 - Belo Horizonte - MG - Brasil
\item
  Laboratório de Botânica Sistemática, Departamento de Botânica e Zoologia, Centro de Biociências, Universidade Federal do Rio Grande do Norte.
\item
  Programa de pós graduação em Sistemática e Evolução, Universidade Federal do Rio Grande do Norte, Natal, Brasil
\item
  Gothenburg Global Biodiversity Centre, University of Gothenburg, Box 461, 405 30 Gothenburg, Sweden
\item
  Department for Biological and Environmental Sciences, University of Gothenburg, Box 461, 405 30 Gothenburg, Sweden
\item
  Royal Botanic Gardens Kew, TW9 3AE, Richmond, Surrey, United Kingdom
\end{enumerate}

@@

Automatic flagging and subsequent filtering are a scalable and reproducible means to identify potentially problematic records in `big data' datasets of species geographic occurrences from public databases (e.g., www.gbif.org). However, it is unclear how much data is lost by filtering is, if similar test should be applied across taxonomic groups, and what is the effect of filtering for downstream analyses. Here, we obtained occurrence records for 16 taxa including plants and animals (terrestrial and marine groups) from www.gbif.org and documented the effect of 13 recently proposed filters in each group. We evaluated the effect of filtering on the inference of species richness patterns and the accuracy of automated conservation assessments as examples for downstream analyses. We found 29-67\% of the records potentially problematic across taxonomic groups. The tests for duplicated information, collection year, basis of record as well as urban areas and political centroids were most important. While many of the flagged records might not be erroneous per se, they might be overly imprecise, increase uncertainty of downstream analyses and generate unnecessary computational burden. Overall, our results stress the importance of (meta-)data exploration for biogeographic research and suggest that automated flagging can help to identify problematic records, but also highlight the necessity to customize tests and thresholds to the taxonomic group and geographic area of interest.

@@

Automated cleaning, automated conservation assessment, data quality, GBIF, Neotropics

@@

Publicly available species distribution data have become a crucial resource in biogeographic research. In particular, the availability of digitized collections from museums and herbaria, and citizen science observations has increased in the last years. As of today, the largest public aggregator for geo-referenced species occurrences data, the Global Biodiversity Information Facility (www.gbif.org), provides access to more than 1.3 billion geo-referenced occurrence records for species from across the globe and the tree of life.

A central challenge to the use of these publicly available species occurrence data in research are erroneous geographic coordinates (Anderson et al., 2016). Errors mostly arise, because public databases integrate records collected with different methodology in different places at different times, often with little centralized curation and meta-data. For instance, erroneous coordinates caused by data-entry errors or automated geo-referencing from vague locality descriptions are common (Maldonado et al., 2015; Yesson et al., 2007) and cause recurrent problems such as records of terrestrial species in the sea, records with coordinates assigned to the centroids of political entities, or records from species in captivity (Zizka, Silvestro, et al., 2019).

Manual data cleaning based on expert knowledge can detect these issues, but is only feasible on small taxonomic or geographic scales and is time-consuming and difficult to reproduce. Automated flagging methods to identify potentially problematic records have been proposed as a scalable alternative for datasets with millions of records from many different taxa. Those methods are usually based on geographic gazetteers (e.g., Chamberlain, 2016; Jin \& Yang, 2020; Zizka, Silvestro, et al., 2019) or on additional data, such as environmental variables (Robertson, Visser, \& Hui, 2016). Additionally, filtering procedures based on record meta-data, such as collection year, record type, and coordinate precisions have been proposed to improve the quality of publicly available occurrence records for biogeographic research (Zizka, Silvestro, et al., 2019). While these methods can identify common errors, it is unclear if errors are equally common across taxonomic groups, and which errors are most severe for downstream analyses.

Erroneous records are especially problematic in conservation, where stakes are high. Recently, automated conservation assessment methods have been developed to use publicly available occurrence records for conservation, in particular to support the assessment procedures for the global Red List of the International Union for the Conservation of Nature (IUCN) (Bachman et al., 2011; Dauby et al., 2017; Pelletier et al., 2018). These methods approximate species' range size, namely the Extent of Occurrence (EOO, which is the area of a convex hull polygon comprising all records of a species), the Area of Occupancy (AOO, which is the sum of the area actually occupied by a species, calculated based on a small-scale regular grid), and the number of locations for a preliminary conservation assessment following IUCN Criterion B. Automated assessments using these methods have been used to speed up global (Stévart et al., 2019; Zizka, Azevedo, et al., 2019) and regional (Cosiaux et al., 2018; Schmidt et al., 2017) Red List assessments. Since especially EOO is sensitive to individual records with erroneous coordinates automated conservation assessments might be prone to bias by erroneous records, especially, if the number of records is low, as is the case for many species from the tropical regions on Earth.

Here, we test the effect of automated filtering of species geographic occurrence records on the number of records available in different plant and animal groups. Furthermore, we test the impact of automated filtering procedures on the accuracy of automated conservation assessments compared to full IUCN assessments. Specifically, we evaluate a pipeline of 13 automated filters to flag possibly erroneous records 1) using record meta-data and 2) using geographic gazetteers. We address thee questions:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Which is the most relevant filter, i.e.~which type of error is most common?
\item
  Does the importance of individual filters differ among taxonomic groups?
\item
  Does automated filtering improve the accuracy of automated conservation assessment compared to the global IUCN Red List?
\end{enumerate}

@@

We did all analysis in the R computing environment (R Core Team, 2019), using standard libraries for data handling and visualization (Garnier, 2018; Hijmans, 2019; Ooms, 2014, 2019; Wickham, 2018). All analysis scripts are available from \url{https://github.com/idiv-biodiversity/effects_of_automated_cleaning}, or from the Appendix 1 in the electronic supplement of this study.

@@

This study is the outcome of a workshop held at the Federal University of Rio Grande do Norte in Natal, Brazil in October 2018 gathering taxonomists and biogeographers working with different taxonomic groups of plants and animals across the Neotropics. Each participant analysed geographic occurrence data from their taxonomic group of interest and provided an assessment of the results for this group. Hence, we chose the study groups based on the expertise of the participants which enabled us to compare the results across the tree of life (at the cost of an arbitrary choice of taxa and taxonomic ranks). We obtained public occurrence records for 16 taxa, including one plant family, eight plant genera, one genus of fungi, two families and one genus of terrestrial arthropods, one family of snakes, one family of skates and one genus of fish (Table 1).

@@

We downloaded occurrence information for all study groups from www.gbif.org using the \texttt{rgbif} v1.4.0 package (Chamberlain, 2017) in R (GBIF.org, 2019a, 2019b, 2019c, 2019d, 2019e, 2019f, 2019g, 2019h, 2019i, 2019j, 2019k, 2019l, 2019m, 2019n, 2019o, 2019p). We chose www.gbif.org for this study, since it is the largest and most widely used provider of species occurrence data and covers the entire tree of life. We downloaded only records with geographic coordinates, and limited the study area to latitudes smaller than 33\(^\circ\) N and longitudes smaller than 35\(^\circ\) W and larger than 120\(^\circ\) W reflecting the Neotropics, our area of expertise (Morrone, 2014).All included taxa are restricted within the Neotropics (natural populations) except Arhynchobatidae, Iridaceae and Neanuridae for which only the Neotropical occurrence was explored.

@@

We followed the cleaning pipeline outlined by Zizka, Silvestro, et al. (2019) and first filtered the raw data downloaded from GBIF (``raw'', hereafter) using meta-data for those records for which it was available, removing: 1) records with a coordinate precision below 100 km (as this represent the grain size of many macro-ecological analyses); 2) fossil records and records of unknown source; 3) records collected before 1945 (before the end of the Second World War, since coordinates of old records are often imprecise and often reflect different land-use types); and 4) records with less than 1 and more than 99 individual records. Furthermore, we rounded the geographic coordinates to four decimal precision and retained only one record per species per location (i.e.~test for duplicated records. In a second step, we used the \texttt{clean\_coordinates} function of the \texttt{CoordinateCleaner\ v2.0-11} package (Zizka, Silvestro, et al., 2019) with default options to flag errors that are common to biological collections, including: sea coordinates, zero coordinates, coordinates assigned to country and province centroids, coordinates within urban areas, and coordinates assigned to biodiversity institutions (``filtered'', hereafter). See Table 2 for a summary of all flagging tests we used.

@@

We first generated species richness maps using 100x100 km grid cells for the raw and filtered dataset respectively, using the \texttt{speciesgeocodeR\ v2.0-10} package (Töpel et al., 2016) and then performed a preliminary conservation assessment for all study groups based on both datasets using the \texttt{ConR\ v1.2.4} package (Dauby et al., 2017). \texttt{ConR} estimates the EOO, AOO, and the number of locations, and then suggests a preliminary conservation status based on Criterion B of the global IUCN Red List. While these assessments are preliminary (see IUCN Standards and Petitions Subcommittee, 2017), they can be a proxy used by the IUCN to speed up full assessments. We then benchmarked the preliminary conservation assessments with the global IUCN Red List assessments, which we obtained from www.iucn.org via the \texttt{rredlist\ v.0.5.0} package (Chamberlain, 2018).

\hypertarget{evaluation-of-the-results}{%
\subsection{Evaluation of the results}\label{evaluation-of-the-results}}

Each specialist evaluated the performance of the raw and cleaned datasets, respectively concerning: 1) the number of flagged records; 2) the importance of specific tests; 3) the accuracy of the overall species richness maps; 4) the accuracy of the preliminary conservation assessment.

@@

We found a total of 215,262 occurrence records, with a median of 3,460 records per study group and 11 records per species (Table 3, Appendix 2). We found most records for Dipsadidae (64,249) and fewest for \emph{Thozetella} (51). The species with most records was \emph{Harengula jaguana} (19,878).

Our automated tests flagged a total of 88,648 records (Fig. \ref{fig:total}), with a median of 40.3\% per group. Overall, the most important test was for duplicated records (on average 32.9\% per taxonomic group). The most important filtering steps based on record meta-data were for the basis of records (2.1\%) and the collection year (3.3\%). The most important automated tests were for urban area (5\%), the capital coordinates (1.1\%) and the centroids of political units (0.5\%, Table 3, see Supplementary material 1 for the absolute numbers). Only few records were flagged by the zero coordinates and biodiversity institutions tests (Fig. \ref{fig:split}), and no records were flagged by the sea test.

\emph{Prosthechea}, \emph{Harengula}, and \emph{Thozetella} had the highest fraction of flagged records (Table 3). In general, the different filters we tested were of similar importance for different study groups. There were few outstanding exceptions, including the particularly high proportions of records flagged by the basis of record test for \emph{Tityus} (7.0\%), Dipsadidae (5.6\%), \emph{Prosthechea} (5.0\%) and \emph{Tillandsia} (4.9\%), by the collection year for Dipsadidae (11.3\%), by the taxonomic identification level for \emph{Tityus} (1.6\%), by the capital coordinates for \emph{Oocephalus} (6.1\%) and \emph{Gaylussacia} (3.2\%), as well as by the urban areas test for \emph{Oocephalus} (13.3\%) and Iridaceae (12.3\%).

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{lmodern}
\usepackage{setspace}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={No one-size-fits-all solution to clean GBIF},
  pdfauthor={Alexander Zizka1,2,; Fernanda Antunes Carvalho4,; Alice Calvente4,; Ana Paula Alves da Silva4,; Andressa Cabral5,; Arthur de Souza Soares4,; Eduardo Calisto Tomaz4,; Fernanda Gondim Lambert Moreira4,; Filipe Cabreirinha Serrano6,; Jéssica Fernanda Ramos Coelho4,; Mabel Rocio Baez-Lizarazo,7; Mariana Ramos Fantinati4,; Matheus Colli-Silva5,; Moabe Ferreira Fernandes8,; Nathália Michelly da Cunha Santos4,; Paolla Gabryelle Cavalcante de Souza4,; Renata Clicia dos Santos‐Costa4,; Thais Ferreira-Araújo4,; Tiago Andrade Borges Santos9,; Tiago Luiz Vieira9,; Valéria Fonseca Vale4,; Alexandre Antonelli10,11,12},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs}
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{caption}
\usepackage{fancyhdr}
\usepackage{booktabs}
\pagestyle{fancy}
\fancyhead[L]{Automated filter effect}
\fancyhead[LE,RO]{\thepage}
\fancyfoot{}
\usepackage{float}
\floatplacement{figure}{H}
\usepackage{lineno}
\linenumbers
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\title{No one-size-fits-all solution to clean GBIF}
\author{Alexander Zizka\textsuperscript{1,2}, \and Fernanda Antunes Carvalho\textsuperscript{4}, \and Alice Calvente\textsuperscript{4}, \and Ana Paula Alves da Silva\textsuperscript{4}, \and Andressa Cabral\textsuperscript{5}, \and Arthur de Souza Soares\textsuperscript{4}, \and Eduardo Calisto Tomaz\textsuperscript{4}, \and Fernanda Gondim Lambert Moreira\textsuperscript{4}, \and Filipe Cabreirinha Serrano\textsuperscript{6}, \and Jéssica Fernanda Ramos Coelho\textsuperscript{4}, \and Mabel Rocio Baez-Lizarazo,\textsuperscript{7} \and Mariana Ramos Fantinati\textsuperscript{4}, \and Matheus Colli-Silva\textsuperscript{5}, \and Moabe Ferreira Fernandes\textsuperscript{8}, \and Nathália Michelly da Cunha Santos\textsuperscript{4}, \and Paolla Gabryelle Cavalcante de Souza\textsuperscript{4}, \and Renata Clicia dos Santos‐Costa\textsuperscript{4}, \and Thais Ferreira-Araújo\textsuperscript{4}, \and Tiago Andrade Borges Santos\textsuperscript{9}, \and Tiago Luiz Vieira\textsuperscript{9}, \and Valéria Fonseca Vale\textsuperscript{4}, \and Alexandre Antonelli\textsuperscript{10,11,12}}
\date{}

@@

\setstretch{1.3}
\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  sDiv, German Center for Integrative Biodiversity Research (iDiv) Halle-Jena-Leipzig, Leipzig, Germany
\item
  Naturalis Biodiversity Center, Leiden, The Netherlands
\item
  Departamento de Genética, Ecologia e Evolução, Universidade Federal de Minas Gerais, Belo Horizonte, Brazil
\item
  Departamento de Botânica e Zoologia, Universidade Federal do Rio Grande do Norte, Natal, Brazil
\item
  Departamento de Botânica, Universidade de São Paulo, São Paulo, Brazil
\item
  Departamento de Ecologia, Universidade de São Paulo, São Paulo, Brazil
\item
  Universidade Federal do Rio Grande do Sul, Porto Alegre, Brazil
\item
  Departamento de Ciências Biológicas, Universidade Estadual de Feira de Santana, Feira de Santana, Brazil
\item
  Universidade Estadual de Feira de Santana, Bahia, Brazil
\item
  Gothenburg Global Biodiversity Centre, University of Gothenburg, Gothenburg, Sweden
\item
  Department for Biological and Environmental Sciences, University of Gothenburg, Gothenburg, Sweden
\item
  Royal Botanic Gardens Kew, Richmond, United Kingdom
\end{enumerate}

@@

Species occurrence records provide the basis for many biodiversity studies. They derive from geo-referenced specimens deposited in natural history collections and visual observations, such as those obtained through various mobile applications. Given the rapid increase in availability of such data, the control of quality and accuracy constitutes a particular concern. Automatic flagging and filtering are a scalable and reproducible means to identify potentially problematic records in datasets from public databases such as the Global Biodiversity Information Facility (GBIF; www.gbif.org). However, it is unclear how much data may be lost by filtering, whether the same tests should be applied across all taxonomic groups, and what is the effect of filtering for common downstream analyses. Here, we evaluate the effect of 13 recently proposed filters on the inference of species richness patterns and automated conservation assessments for 16 Neotropical taxa including plants and animals, terrestrial and marine, downloaded from GBIF. We find that 29-67\% of the records are potentially erroneous, with large variation across taxonomic groups. Tests for duplicated information, collection year, basis of record as well as urban areas and political centroids have the greatest effect. While many flagged records might not be \emph{de facto} erroneous, they could be overly imprecise and increase uncertainty in downstream analyses. Automated flagging can help in identifying problematic records, but requires customization of which tests and thresholds should be applied to the taxonomic group and geographic area under focus. Our results stress the importance of thorough exploration of the meta-data associated with species records for biodiversity research.

@@

Automated cleaning, automated conservation assessment, data quality, GBIF, Neotropics.

@@

Publicly available species distribution data have become a crucial resource in biodiversity research, including studies in ecology, biogeography, systematics and conservation biology. In particular, the availability of digitized collections from museums and herbaria, and citizen science observations has increased drastically over the last few years. As of today, the largest public aggregator for geo-referenced species occurrences data, the Global Biodiversity Information Facility (www.gbif.org), provides access to more than 1.3 billion geo-referenced occurrence records for species from across the globe and the tree of life.

A central challenge to the use of these publicly available species occurrence data in research are erroneous geographic coordinates (Anderson et al., 2016). Errors mostly arise because public databases integrate records collected with different methodologies in different places, at different times; often without centralized curation and only rudimentary meta-data. For instance, erroneous coordinates caused by data-entry errors or automated geo-referencing from vague locality descriptions are common (Maldonado et al., 2015; Yesson et al., 2007) and cause recurrent problems such as records of terrestrial species in the sea, records with coordinates assigned to the centroids of political entities, or records of species in captivity (Zizka, Silvestro, et al., 2019).

Manual data cleaning based on expert knowledge can detect these issues, but it is only feasible on small taxonomic or geographic scales, and it is time-consuming and difficult to reproduce. As an alternative, automated flagging methods to identify potentially problematic records have been proposed as a scalable option, as they are able to deal with datasets containing up to millions of records and many different taxa. Those methods are usually based on geographic gazetteers (e.g., Chamberlain, 2016; Jin \& Yang, 2020; Zizka, Silvestro, et al., 2019) or on additional data, such as environmental variables (Robertson, Visser, \& Hui, 2016). Additionally, filtering procedures based on record meta-data, such as collection year, record type, and coordinate precisions have been proposed to improve the quality of publicly available occurrence records for biodiveristy research (Zizka, Silvestro, et al., 2019).

Erroneous records are especially problematic in conservation, where stakes are high. Recently proposed methods for automated conservation assessments could support the formal assessment procedures for the global Red List of the International Union for the Conservation of Nature (IUCN) (Bachman et al., 2011; Dauby et al., 2017; Pelletier et al., 2018). These methods approximate species' range size, namely the Extent of Occurrence (EOO, which is the area of a convex hull polygon comprising all records of a species), the Area of Occupancy (AOO, which is the sum of the area actually occupied by a species, calculated based on a small-scale regular grid), and the number of locations for a preliminary conservation assessment following IUCN Criterion B (``Geographic range''). These methods have been used topropose preliminary global (Stévart et al., 2019; Zizka, Azevedo, et al., 2019) and regional (Cosiaux et al., 2018; Schmidt et al., 2017) Red List assessments. However, all metrics but especially EOO, are sensitive to individual records with erroneous coordinates \textbf{(REFERENCE)}. Automated conservation assessments may therefore be biased, particularly if the number of records are few, as it is the case for many tropical species.

While newly proposed methods for automated cleaning of species records hold great promise for biodiversity research, their use across taxonomic groups and datasets remains poorly explored. Here, we test the effect of automated filtering of species geographic occurrence records on the number of records available in different groups of animals, funghi, and plants. Furthermore, we test the impact of automated filtering procedures for the accuracy of automated conservation assessments compared to full IUCN assessments. Specifically, we evaluate a pipeline of 13 automated filters to flag possibly erroneous records by using 1) record meta-data and 2) geographic gazetteers. We address three questions:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Which filters lead to the biggest loss of data when applied?
\item
  Does the importance of individual filters differ among taxonomic groups?
\item
  Does automated filtering improve the accuracy of automated conservation assessments?
\end{enumerate}

@@

@@

This study is the outcome of a workshop held at the Federal University of Rio Grande do Norte in Natal, Brazil in October 2018 gathering graduate students, taxonomists and biogeographers working with different taxonomic groups of animals, funghi, and plants across the Neotropics. Each participant analysed geographic occurrence data from their taxonomic group of interest and provided an informed assessment of the results for their group. Hence, we include groups based on the expertise of the participants wrather than following some arbitrary choice of taxa and taxonomic ranks. We acknowledge a varying degree in documented expertise and number of years working on each group. We obtained public occurrence records for 16 taxa, including one plant family, eight plant genera, one genus of fungi, two families and one genus of terrestrial arthropods, one family of snakes, one family of skates, and one genus of bony fish (Table 1).

@@

We downloaded occurrence information for all study groups from www.gbif.org using the \texttt{rgbif} v1.4.0 package (Chamberlain, 2017) in R (GBIF.org, 2019a-p). We downloaded only records with geographic coordinates and limited the study area to latitudes smaller than 33\(^\circ\) N and longitudes smaller than 35\(^\circ\) W and larger than 120\(^\circ\) W reflecting the Neotropics, our main area of expertise (Morrone, 2014). The natural distribution of all included taxa are confined to the Neotropics except Arhynchobatidae, Diogenidae, Dipsadidae, \emph{Gaylussacia}, Iridaceae, Neanuridae, and \emph{Tillandsia} for which we only obtained the Neotropical occurrences.

@@

We followed the cleaning pipeline outlined by Zizka, Silvestro, et al. (2019) and first filtered the raw data downloaded from GBIF (``raw'', hereafter) using meta-data for those records for which they were available, removing: 1) records with a coordinate precision below 100 km (as this represent the grain size of many macro-ecological analyses); 2) fossil records and records of unknown source; 3) records collected before 1945 (before the end of the Second World War, since coordinates of old records are often imprecise); and 4) records with an individual count of less than one and more than 99. Furthermore, we rounded the geographic coordinates to four decimal precision and retained only one record per species per location (i.e.~test for duplicated records). In a second step, we used the \texttt{clean\_coordinates} function of the \texttt{CoordinateCleaner\ v2.0-11} package (Zizka, Silvestro, et al., 2019) with default options to flag errors that are common to biological data sets (``filtered'', hereafter). These including: coordinates in the sea for terrestrial taxa, coordinates containing only zeros, coordinates assigned to country and province centroids, coordinates within urban areas, and coordinates assigned to biodiversity institutions. See Table 2 for a summary of all flagging tests we used.

@@

We first generated species richness maps using 100x100 km grid cells for the raw and filtered datasets respectively, using the package \texttt{speciesgeocodeR\ v2.0-10} (Töpel et al., 2016). We then performed an automated conservation assessment for all study groups based on both datasets using the \texttt{ConR\ v1.2.4} package (Dauby et al., 2017). \texttt{ConR} estimates the EOO, AOO, and the number of locations, and then suggests a preliminary conservation status based on Criterion B of the global IUCN Red List. While these assessments are preliminary (see IUCN Standards and Petitions Subcommittee, 2017), they can be a proxy used by the IUCN to speed up full assessments. We then benchmarked the preliminary conservation assessments against the global IUCN Red List assessments as far as available for the same taxa, which we obtained from www.iucn.org via the \texttt{rredlist\ v.0.5.0} package (Chamberlain, 2018).

\hypertarget{evaluation-of-results}{%
\subsection{Evaluation of results}\label{evaluation-of-results}}

Each specialist provided an informed assessment on the performance of the raw and cleaned datasets, respectively concerning: 1) the number of flagged records; 2) the importance of specific tests; 3) the accuracy of the overall species richness maps; 4) the accuracy of the preliminary conservation assessments.

We did all analyses in the R computing environment (R Core Team, 2019), using standard libraries for data handling and visualization (Garnier, 2018; Hijmans, 2019; Ooms, 2014, 2019; Wickham, 2018). All scripts are available from \url{https://github.com/idiv-biodiversity/effects_of_automated_cleaning}, or from the Appendix 1 in the electronic supplement of this study.

@@

We retrieved a total of 215,262 species occurrence records, with a median of 3,460 records per study group and 11 records per species (Table 3, Appendix 2). We obtained most records for Dipsadidae (64,249) and fewest for \emph{Thozetella} (51). The species with most records was \emph{Harengula jaguana} (19,878).

Our automated tests flagged a total of 88,648 records (Fig. \ref{fig:total}), with a median of 40.3\% per group. Overall, the most important test was for duplicated records (on average 32.9\% per taxonomic group). The filtering steps based on record meta-data that flagged the largest number of records were the basis of records (2.1\%) and the collection year (3.3\%). The most important automated tests were for urban area (5\%), the capital coordinates (1.1\%) and the centroids of political units (0.5\%, see Table 3 and Supplementary material 1 for further details and the absolute numbers). Only few records were flagged by the zero coordinates and biodiversity institution tests (Fig. \ref{fig:split}), and no records were flagged by the sea test.

\emph{Prosthechea}, \emph{Harengula}, and \emph{Thozetella} had the highest fraction of flagged records (Table 3). In general, the different filters we tested were of similar importance for different study groups. There were few outstanding exceptions, including the particularly high proportions of records flagged on the ``basis of record test'' for \emph{Tityus} (7.0\%), Dipsadidae (5.6\%), \emph{Prosthechea} (5.0\%) and \emph{Tillandsia} (4.9\%), by the collection year for Dipsadidae (11.3\%), by the taxonomic identification level for \emph{Tityus} (1.6\%), by the capital coordinates for \emph{Oocephalus} (6.1\%) and \emph{Gaylussacia} (3.2\%), and by the urban areas test for \emph{Oocephalus} (13.3\%) and Iridaceae (12.3\%).

Geographically, the records flagged by the ``basis of record'' and ``individual count'' tests were concentrated in Central America and southern North America, and a relatively high number of records flagged due to their proximity to the centroids of political entities were located on Caribbean islands (Fig. \ref{fig:split}). See Supplementary material 2 for species richness maps using the raw and cleaned data for all study groups.
